---
title: "Untitled"
output: html_document
---
## 8.3

### a

```{r}
# Load data
library(dplyr)
library(tidyr)
library(ggplot2)
schools.list = lapply(1:8, function(i) {
  s.tbl = paste0('http://www.stat.washington.edu/people/pdhoff/Book/Data/hwdata/school', i, '.dat') %>%
    url %>%
    read.table
  
  data.frame(
    school = i,
    hours = s.tbl[, 1] %>% as.numeric
  )
})
schools.raw = do.call(rbind, schools.list)
Y = schools.raw
# Prior
mu0 = 7
g20 = 5
t20 = 10
eta0 = 2
s20 = 15
nu0 = 2
# Number of schools. Y[, 1] are school ids
m = length(unique(Y[, 1]))
# Starting values - use sample mean and variance
n = sv = ybar = rep(NA, m)
for (j in 1:m) {
  Y_j = Y[Y[, 1] == j, 2]
  ybar[j] = mean(Y_j)
  sv[j] = var(Y_j)
  n[j] = length(Y_j)
}
# Let initial theta estimates be the sample means
# Similarly, let initial values of sigma2, mu, and tau2 be "sample mean and
# variance"
theta = ybar
sigma2 = mean(sv)
mu = mean(theta)
tau2 = var(theta)
# MCMC
S = 1500
THETA = matrix(nrow = S, ncol = m)
# Storing sigma, mu, theta together
SMT = matrix(nrow = S, ncol = 3)
colnames(SMT) = c('sigma2', 'mu', 'tau2')
for (s in 1:S) {
  # Sample thetas
  for (j in 1:m) {
    vtheta = 1 / (n[j] / sigma2 + 1 / tau2)
    etheta = vtheta * (ybar[j] * n[j] / sigma2 + mu / tau2)
    theta[j] = rnorm(1, etheta, sqrt(vtheta))
  }
  
  # Sample sigma2
  nun = nu0 + sum(n) # TODO: Could cache this
  ss = nu0 * s20
  # Pool variance
  for (j in 1:m) {
    ss = ss + sum((Y[Y[, 1] == j, 2] - theta[j])^2)
  }
  sigma2 = 1 / rgamma(1, nun / 2, ss / 2)
  
  # Sample mu
  vmu = 1 / (m / tau2 + 1 /g20)
  emu = vmu * (m * mean(theta) / tau2 + mu0 / g20)
  mu = rnorm(1, emu, sqrt(vmu))
  
  # Sample tau2
  etam = eta0 + m
  ss = eta0 * t20 + sum((theta - mu)^2)
  tau2 = 1 / rgamma(1, etam / 2, ss / 2)
  
  # Store params
  THETA[s, ] = theta
  SMT[s, ] = c(sigma2, mu, tau2)
}
```

Assess convergence with diagnostic boxplots:

```{r echo=FALSE}
smt.df = data.frame(SMT)
colnames(smt.df) = c('sigma2', 'mu', 'tau2')
smt.df$s = 1:S
cut_size = 10
smt.df = smt.df %>%
  tbl_df %>%
  mutate(scut = cut(s, breaks = cut_size)) %>%
  gather('variable', 'value', sigma2:tau2)
ggplot(smt.df, aes(x = scut, y = value)) +
  facet_wrap(~ variable, scales = 'free_y') +
  geom_boxplot() +
  theme(axis.text.x = element_blank()) +
  xlab('Samples')
```

Evaluate effective sample size:

```{r}
# Tweak number of samples until all of the below are above 1000
library(coda)
effectiveSize(SMT[, 1])
effectiveSize(SMT[, 2])
effectiveSize(SMT[, 3])
```

### b

Posterior means and confidence intervals

```{r}
t(apply(SMT, MARGIN = 2, FUN = quantile, probs = c(0.025, 0.5, 0.975)))
```

Comparing posterior to prior:
```{r}
# For dinvgamma
library(MCMCpack)
sigma2_prior = data.frame(
  value = seq(10, 22.5, by = 0.1),
  density = dinvgamma(seq(10, 22.5, by = 0.1), nu0 / 2, nu0 * s20 / 2),
  variable = 'sigma2'
)
tau2_prior = data.frame(
  value = seq(0, 30, by = 0.1),
  density = dinvgamma(seq(0, 30, by = 0.1), eta0 / 2, eta0 * t20 / 2),
  variable = 'tau2'
)
mu_prior = data.frame(
  value = seq(0, 12, by = 0.1),
  density = dnorm(seq(0, 12, by = 0.1), mu0, sqrt(g20)),
  variable = 'mu'
)
priors = rbind(sigma2_prior, tau2_prior, mu_prior)
priors$dist = 'prior'
smt.df$dist = 'posterior'
ggplot(priors, aes(x = value, y = density, color = dist)) +
  geom_line() +
  geom_density(data = smt.df, mapping = aes(x = value, y = ..density..)) +
  facet_wrap(~ variable, scales = 'free')
```

Our prior estimates for $\mu$ and $\tau^2$ were fairly estimate, but our 
estimate for $\sigma^2$ was very far off. After this analysis, we have estimates
for $\mu$, the average amount of hours of schoolwork spent at a typical school, 
$\tau^2$, the variability between schools in the average hours of schoolwork, 
and $\sigma^2$, the variability among students' hours in each school.

### c
```{r}
t20_prior = (1 / rgamma(1e6, eta0 / 2, eta0 * t20 / 2))
s20_prior = (1 / rgamma(1e6, nu0 / 2, nu0 * s20 / 2))
R_prior = data.frame(
  value = (t20_prior) / (t20_prior + s20_prior),
  dist = 'prior'
)
R_post = data.frame(
  value = SMT[, 'tau2'] / (SMT[, 'tau2'] + SMT[, 'sigma2']),
  dist = 'posterior'
)
ggplot(R_prior, aes(x = value, y = ..density.., color = dist)) +
  geom_density(data = R_prior) +
  geom_density(data = R_post)
mean(R_post$value)
```

$R$ measures how much of the total variance in our data is between-group. Our
prior didn't contain much information about this quantity, but after inference,
we expect that around 25\% of our variance comes from between group variance
($\tau^2$).

### d

```{r}
theta7_lt_6 = THETA[, 7] < THETA[, 6]
mean(theta7_lt_6)
theta7_smallest = (THETA[, 7] < THETA[, -7]) %>%
  apply(MARGIN = 1, FUN = all)
mean(theta7_smallest)
```

### e


```{r}
relationship = data.frame(
  sample_average = ybar,
  post_exp = colMeans(THETA),
  school = 1:length(ybar)
)
ggplot(relationship, aes(x = sample_average, y = post_exp, label = school)) +
  geom_text() +
  geom_abline(slope = 1, intercept = 0) +
  geom_hline(yintercept = mean(schools.raw[, 'hours']), lty = 2) +
  annotate('text', x = 10, y = 7.9, label = paste0("Pooled sample mean ", round(mean(schools.raw[, 'hours']), 2))) +
  geom_hline(yintercept = mean(SMT[, 'mu']), color = 'red') +
  annotate('text', x = 10, y = 7.4, label = paste0("Posterior exp. mu ", round(mean(SMT[, 'mu']), 2)), color = 'red')
```